---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---


## Loading and preprocessing the data  

```{r}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl,destfile="./activity.zip",method="curl")
unzip(zipfile="./activity.zip",exdir=".")

df<-read.csv("activity.csv", na.strings = "NA")
df$date<-as.Date(df$date)
str(df)
```


## What is mean total number of steps taken per day?
### I create a summary table with mean, median and total number of steps per day  

```{r}
library(dplyr)
totsteps<-group_by(df, date)
sumsteps<-summarize(totsteps, steps=mean(steps, na.rm=T))
sumsteps<-rename(sumsteps, Mean_steps=steps)
sumsteps<-merge(sumsteps, summarize(totsteps, steps=median(steps, na.rm=T)))
sumsteps<-rename(sumsteps, Median_steps=steps)
sumsteps<-merge(sumsteps, summarize(totsteps, steps=sum(steps, na.rm=T)))
sumsteps<-rename(sumsteps, Sum_steps=steps)
str(sumsteps)

```
  
### I create a histogram of the total number of steps taken each day  

```{r}
hist(sumsteps$Sum_steps, main = "Total number of steps per day", xlab = "Total number of steps in a day")
```

### I report a table with mean and median values of steps per day  

```{r}
subset(sumsteps, select=c(date, Mean_steps, Median_steps))
```


## What is the average daily activity pattern?

### I group the number of steps per interval and average them

```{r}
pattern<-group_by(df, interval)
pattern<-summarize(pattern, steps=mean(steps, na.rm=T))
str(pattern)
```


### I create a time series plot of the x-minute interval and the average number of steps acen across all days

```{r}
with(pattern, plot(interval, steps, type="l", main="Average daily activity pattern", xlab="5-minute interval", ylab="Mean number of steps"))
```



## Imputing missing values

### I report the number of rows with missing values in the dataset

```{r}
nrow(df[!complete.cases(df),])
```
### I impute the missing values as the mean of the average values of steps in that day and in that interval; I will check the number of incomplete cases in the new dataframe.

```{r}

imp<-df

for (i in pattern$interval) {
    imp[imp$interval == i & is.na(imp$steps), ]$steps <-         pattern$steps[pattern$interval == i]
}
print(paste("The number of incomplete rows is now ", nrow(imp[!complete.cases(imp),])))

```
### I create a new summary table with imputed data.

```{r}
totimpute<-group_by(imp, date)
sum2<-summarize(totimpute, steps=mean(steps, na.rm=T))
sum2<-rename(sum2, Mean_steps=steps)
sum2<-merge(sum2, summarize(totimpute, steps=median(steps, na.rm=T)))
sum2<-rename(sum2, Median_steps=steps)
sum2<-merge(sum2, summarize(totimpute, steps=sum(steps, na.rm=T)))
sum2<-rename(sum2, Sum_steps=steps)
str(sum2)
```
### I create a histogram of the total number of steps taken each day with imputed data

```{r}
hist(sum2$Sum_steps, main = "Total number of steps per day", xlab = "Total number of steps in a day")
```
### I report a table with mean and median values of steps per day and the differences with the previous values

```{r}
delta<-subset(sum2, select=c(date, Mean_steps, Median_steps))
delta$difference_Mean<-  sum2$Mean_steps + sumsteps$Mean_steps*(-1)
delta$difference_Median<- sum2$Median_steps + sumsteps$Median_steps*(-1)
```

### Except for the days that had no measurements at all, there are no differences in mean and median values per day

## Are there differences in activity patterns between weekdays and weekends?

### I add a new column to the imputed dataset with the weekday and then I make a panel plot with the time series of the 5-minute interval and the average number of steps taken, averaged across all weekdays or weekends.
```{r}
imp$Weekday<-weekdays(imp$date)
imp[imp$Weekday == "lunedì" | imp$Weekday == "martedì" | imp$Weekday == "mercoledì" | imp$Weekday == "giovedì" | imp$Weekday == "venerdì", ]$Weekday <- "weekday"
imp[imp$Weekday == "sabato" | imp$Weekday == "domenica", ] $Weekday<-"weekend" 

imp$Weekday <- factor(imp$Weekday)

weekend <- summarise(group_by(imp, interval)[imp$Weekday=="weekend", ], steps=mean(steps))
weekend$day <- "weekend"

weekday <- summarise(group_by(imp, interval)[imp$Weekday=="weekday", ], steps=mean(steps))
weekday$day <- "weekday"

week<-rbind(weekend, weekday)


library(lattice)
library(png)
xyplot(steps~interval|day, week, layout=c(1,2), type="l", main="Average Daily Pattern: weekend vs weekday")



```

